Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-02-21 paper/paper.tex
pass

The SYN Flood attack is a denial-of-service attack that exploits a vulnerability in the TCP/IP handshake
to establish a TCP connection~\cite{cloudfare-synflood}. The handshake
works as follows in normal circumstances. First, a client sends a SYN
packet to the server, requesting a connection. Second, the server
responds with a SYN-ACK packet to the client. Third, the client
responds with an ACK message and the connection is established. Aware
of the protocol, an attacker sends multiple SYN packets to different
ports of a server, often using fake IP addresses. Then, after the
server responds with a SYN-ACK packet, the client keeps sending other
SYN packets to avoid the connection to time out. Without proper
protection, the server accepts these malicious requests and eventually
legitimate requests cannot be satisfied due to resource exhaustion.

Figure~\ref{fig:synflood-example} shows a Suricata rule that captures
this attack. This rule was obtained from the Emerging Threats Open
Ruleset~\cite{emerging-threats-open}. The relevant options in the
rule appear in bold---the option \CodeIn{flags: S,12}, identifies a
SYN packet in a TCP packet and the option \CodeIn{threshold: type
  both, track by\_dst, count 5000, seconds 5} indicates that a high
volume of such packets should be requested in a short period of time.

In the following, we illustrate how \tname{} proceeds for obtaining
the rule for the SYN Flood attack. The challenge in synthesizing rules
is to discover the set of options in a rule that makes the \nids\ to
capture the malicious traffic, but allow the benign traffic to go
through.  The other parts of the rule (\ie{}, the action and the
header) are either inferred or left constant. For example, in this
case, \tname{} sets the action header to \CodeIn{alert} and sets the
header to \CodeIn{<proto> any any -> any any}, indicating that it
infers the target protocol and uses the most general flow description
possible, \ie{}, it instructs the \nids\ to analyze the traffic
flowing from/to any address and port.  We understand that this is
domain-specific information and assume that the system administrator
would know how to customize it.

\tname{} proceeds as follows to synthesize rules for this
attack. First, it analyzes the positive (\ie, malicious) traffic and
extracts options expressed in that traffic. In this case, the
malicious traffic is characterized as a sequence of 5000
messages. \tname{} finds the following set of options at this stage:

@@

\noindent
\tname{} only reports a rule if the \nids\ instantiated with that rule
is able to capture the traffic passed on input. Consequently, by
construction, a rule that uses this set of options captures the
malicious traffic. Unfortunately, this rule is potentially
overspecified. Informally, the rule includes ``too specific''
characteristics of the malicious traffic and could potentially result
in false negatives. For example, in this particular case, note that
the option \CodeIn{window: 512} is not included in the golden rule
from Figure~\ref{fig:synflood-example}.

Second, \tname{} uses negative (\ie{}, benign) traffic to mitigate the
overfitting issue. In this step, \tname{} searches for alternative
rules that preserves the invarianxt that a rule should capture the
positive traffic but not capture the negative traffic. It uses the
rule obtained in the previous step, containing three options, to
bootstrap the search for rules. The search generates rules containing
subsets of the options from the seed rule.

@@

Finally, \tname{} ranks the inferred rules according to heuristics
extracted from a database of existing rules. One example heuristic
function we used to rank the generated rules was \Luc{a quantidade de opcoes na regra}
\Fix{...elaborate...}. The list below shows the rule options for the
top-5 rules \tname{} produces for this attack.

To sum up, \tname{} uses the messages that manifest an attack in the
first step (to create an overspecific seed rule), benign traffic in
the second step (to create a set of plausible rules from the seed
rule), and existing rules from popular sources in the third step (to
rank the generated rules).

\vspace{1ex}
\noindent
\textbf{Differences.}~Note that there are differences between the
options produced in the first step of \tname{} and the options from
the golden rule, as per Figure~\ref{fig:synflood-example}. The option
\CodeIn{flags} does not mention parameters \CodeIn{1 2}, there is an
extra option \CodeIn{window}, and there are differences in the option
\CodeIn{threshold}. Considering the option \CodeIn{flags}, the first
parameter of that option indicates what flags in the packet must be
set while the second parameter indicates non-mandatory flags (\ie{},
``don't cares''). The option \CodeIn{flags: S} indicates that the flag
``S'' must be set and all flags not mentioned (``1'' and ``2'', in our
case) must be unset for the \nids\ to capture the
traffic. Consequently, that choice of option alone could result in
false negatives and false positives. An attack that satisfies all
other conditions in a rule and have flags ``1'' or ``2'' set would be
captured by the original rule but not the inferred rule---a case of
false negative. Likewise, an attack that satisfies all conditions and
have flags ``1'' and ``2'' unset would be captured by the inferred
rule but not the original rule---a case of false positive. The set of
positive examples is to blame in this case. To mitigate this issue,
the input messages should have covered cases with options ``1'' or
``2'' set. Considering option \CodeIn{window: 512}, \tname{} is able
to generate rules without that option, as discussed above. Considering
option \CodeIn{threshold}, the values used in the golden rule are
domain-specific. \tname{} uses default values for these options. We
conjecture that the system administrator would be able to configure
this option.

>>>>

The SYN Flood attack is a denial-of-service attack that exploits a vulnerability in the TCP/IP handshake
to establish a TCP connection~\cite{cloudfare-synflood}. The handshake
works as follows in normal circumstances. First, a client sends a SYN
packet to the server, requesting a connection. Second, the server
responds with a SYN-ACK packet to the client. Third, the client
responds with an ACK message and the connection is established. Aware
of the protocol, an attacker sends multiple SYN packets to different
ports of a server, often using fake IP addresses. Then, after the
server responds with a SYN-ACK packet, the client keeps sending other
SYN packets to avoid the connection to time out. Without proper
protection, the server accepts these malicious requests and eventually
legitimate requests cannot be satisfied due to resource exhaustion.
Figure~\ref{fig:synflood-example} shows a Suricata rule that captures
this attack. This rule was obtained from the Emerging Threats Open
Ruleset~\cite{emerging-threats-open}. The relevant options in the
rule appear in bold---the option \CodeIn{flags: S,12}, identifies a
SYN packet in a TCP packet and the option \CodeIn{threshold: type
  both, track by\_dst, count 5000, seconds 5} indicates that a high
volume of such packets should be requested in a short period of
time. The other options are for documentation purposes.

In the following, we discuss the rule produced by \tname{} to capture
this attack. The action and the header of the rules that \tname{}
creates are either inferred or left constant. For example, in this
case, \tname{} sets the action to \CodeIn{alert} and sets the header
to \CodeIn{<proto> any any -> any any}, indicating that it infers the
target protocol and uses the most general flow description possible,
\ie{}, it instructs the \nids\ to analyze the traffic flowing from/to
any address and port.  We understand that the choice of these
parameters are domain-specific and assume that the sys admin would
know how (and want) to customize them.

The main challenge in synthesizing rules is to discover the set of
options in a rule that blocks the malicious traffic, but allows the
benign traffic to pass. \tname{} proceeds as follows to synthesize
rules for this attack.

First, \tname{} analyzes the malicious traffic and extracts options
expressed in that traffic. Recall from
Section~\ref{sec:rules-and-packets} that rule options can be inferred
from packet contents. In this case, however, that information alone is
insufficient as the malicious traffic is characterized as a sequence
of 5K\Mar{really 5K?}\Mar{...and where did they come from?}
messages. \tname{} finds the following set of options at this stage by
analyzing the flow and contents of the messages:

@@

\noindent
By construction, a rule that uses this set of options captures the
malicious traffic. Unfortunately, this rule is potentially
overspecified. Informally, the rule includes ``too specific''
characteristics of the malicious traffic and could potentially result
in false negatives. For example, in this particular case, note that
the option \CodeIn{window: 512} is not included in the golden rule
from Figure~\ref{fig:synflood-example}.

Second, \tname{} uses benign traffic to mitigate the overfitting
issue. In this step, \tname{} searches for alternative rules that
preserves the invarianxt that a rule should capture the positive
malicious traffic but not capture the negative benign traffic. It uses
the rule obtained in the previous step, containing three options, to
bootstrap the search for alternative rules. The search generates rules
containing subsets of the options from the seed rule.

@@

Finally, \tname{} ranks the inferred rules according to heuristics
learned from a database of existing rules. For example, one heuristic
function \tname{} uses for ranking leverages the amount of options a
rule of a given protocol typically has---the score of a rule on this
function is proportional to the popularity of the number of options it
declares. The list below shows the rule options for the top-5 rules
\tname{} produces for this attack.\Mar{algum problema em fazer isto?
  exemplos concretos ajudam muito!}

To sum up, \tname{} uses the messages that manifest an attack in the
first step to create an overspecific seed rule, benign traffic in the
second step to create a set of plausible rules from the seed rule, and
existing rules from popular sources in the third step to rank
alternative plausible rules.

\vspace{1ex}
\noindent
\textbf{Differences.}~Note that there are differences between the
options produced in the first step of \tname{} and the options from
the golden rule, as per Figure~\ref{fig:synflood-example}. The option
\CodeIn{flags} does not mention parameters \CodeIn{1 2}, there is an
extra option \CodeIn{window}, and there are differences in the option
\CodeIn{threshold}. Considering the option \CodeIn{flags}, the first
parameter of that option indicates what flags in the packet must be
set while the second parameter indicates non-mandatory flags (\ie{},
``don't cares''). More specifically, the option \CodeIn{flags: S}
indicates that the flag ``S'' must be set and all flags not mentioned
(including ``1'' and ``2'') must be unset for the \nids\ to capture
the traffic. Consequently, that choice of option alone could result in
false negatives and false positives. An attack that satisfies all
other conditions in a rule and has flags ``1'' or ``2'' set would be
captured by the original rule but not captured by the inferred rule---a case of
false negative. Likewise, an attack that satisfies all conditions and
have flags ``1'' and ``2'' unset would be captured by the inferred
rule but not captured by the original rule---a case of false positive. The set of
positive examples is to blame in this case. To mitigate this issue,
the input messages should have covered cases with options ``1'' or
``2'' set. Considering option \CodeIn{window: 512}, \tname{} is able
to generate rules without that option, as discussed above. Considering
option \CodeIn{threshold}, the values used in the golden rule are
domain-specific. \tname{} uses default values for these options. We
conjecture that the system administrator would be able to configure
this option.
